Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-16 paper.tex
new title
editing intro.  still in progress so don't criticize yet.  sorry taking so long...

\title{Cooling Requirements for the Vertical Shear Instability in
  astrophysical disks \\ and Implications for Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

@@

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent MHD simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative mechanisms 
to angular momentum transport in PPDs, such as purely hydrodynamic
turbulence,  remain of interest \citep{armitage10}. In addition to
mass accretion, turbulence in PPDs will also
impact dust evolution and planetesimal formation \citep{chiang10}.

One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI). The VSI is an 
axisymmetric linear instability 
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. 
Vertical shear occurs whenever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in disks, except at special locations like the midplane.

The VSI  in disks is fundamentally equivalent to the
Goldreich-Schubert-Fricke instability 
\citep[GSFI,][]{goldreich67,fricke68} in differentially rotating stars.   
A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric, inviscid, neutrally
stratified disk, the exchange of fluid parcels vertically, while
conserving their specific angular momentum, can result in a new
configuration of lower energy if the distribution of specific angular
momentum varies with $z$. The VSI is characterized by perturbations
with large vertical length-scales  relative to radial length-scales.

>>>>

\title{Cooling Requirements for the Vertical Shear Instability 
  in Protoplanetary Disks}
%\title{Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula} 
\author{Min-Kai Lin \& Andrew Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\email{minkailin@email.arizona.edu, youdin@email.arizona.edu}

@@

However, MHD turbulence may not be applicable to all  
accretion disks. For example, there may be regions in protoplanetary  
disks (PPDs) which do not support the MRI 
\citep{gammie96,landry13}. In fact, recent MHD simulations
suggest significant portions of PPDs remain laminar \citep[e.g.][and references
therein]{lesur14,bai15,gressel15}. Thus, alternative mechanisms 
to angular momentum transport in PPDs, such as purely hydrodynamic
turbulence,  remain of interest \citep{armitage10}. In addition to
mass accretion, turbulence in PPDs will also
impact dust evolution and planetesimal formation \citep{yl07, chiang10}.

One possible route to hydrodynamic turbulence is the vertical shear
instability (VSI; citealp{urpin98, nelson13}).  The VSI  in disks is the same
basic mechanism as the Goldreich-Schubert-Fricke instability 
\citep[GSFI,][]{goldreich67,fricke68} in differentially rotating stars.   
The VSI arises when vertical shear, i.e. a variation in orbital motion with 
height, destabilizes inertial-gravity waves, oscillations 
with rotation and buoyancy as restoring forces.  Vertical shear
occurs wherever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in disks, except at special locations like the midplane.

To overcome the stabilizing effects of rotation, the VSI triggers motions which are vertically elongated 
and radially narrow.  This geometry is only minimally affected by the stabilizing radial gradient of angular momentum.
The vertical elongation taps vertical shear, but is also subject to the stabilizing effects of vertical buoyancy.
Vertical buoyancy could be weak if there is strong internal heating, for instance close to the star (refs).  However, 
realistic disk models with external irradiation tend to have significant vertical buoyancy, at last outside $\sim 1 AU$ (refs).

To overcome vertical buoyancy, the VSI requires a short cooling time \citep{goldreich67, nelson13}.  Rapid cooling via radiation
brings a moving fluid into thermal equilibrium with its surroundings diminishing buoyant stabilization.  An isothermal equation
of state implies instantaneous cooling, and is the ideal

The VSI is an axisymmetric linear instability 
in three-dimensional (3D) disks in which the rotation frequency
$\Omega$ depends on the height $z$ from the disk midplane. 
Vertical shear occurs whenever the disk is baroclinic, i.e.\ when constant 
density and constant pressure surfaces are misaligned.  Baroclinicity, and thus vertical shear, is 
practically unavoidable in disks, except at special locations like the midplane.

A height-dependence in $\Omega$, or vertical shear, can render
inertial waves unstable. \cite{barker15} gives a simple 
interpretation of the VSI: in an axisymmetric, inviscid, neutrally
stratified disk, the exchange of fluid parcels vertically, while
conserving their specific angular momentum, can result in a new
configuration of lower energy if the distribution of specific angular
momentum varies with $z$. The VSI is characterized by perturbations
with large vertical length-scales  relative to radial length-scales.
